Enhanced Overall Local Accuracy Framework For Addressing Feature Bias and Data Imbalance in ASD Classification

### 2025-06-05 Paper/BichHao/main.tex
paper Hao

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage[english,vietnamese]{babel}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{array}
\usepackage[table,xcdraw]{xcolor}
\usepackage{placeins}
\usepackage{tcolorbox}
\usepackage{verbatim}
\usepackage{enumitem}
\usepackage{float}
\usepackage{caption}
\usepackage{float}
\usepackage{tabularx}       
\usepackage{booktabs}       
\usepackage{caption}        
\usepackage{float} 
\usepackage{url}
\usepackage{hyperref}
% \usepackage{autoref}
\usepackage{subcaption}

% \addto\captionsvietnamese{
%   \renewcommand{\tablename}{Table}
% }
\addto\captionsvietnamese{
  \renewcommand{\abstractname}{Abstract}
}
\addto\captionsvietnamese{
  \renewcommand{\refname}{References}
}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Overcoming Data Imbalance and Feature Bias in Autism Classification with Overall Local Accuracy }

\author{
Thanh Nguyen Van Quoc $^{1}$, Hao Nguyen Thi Bich $^{2}$, Nhut Nguyen Minh$^{3}$, Thuan Nguyen Dinh$^{4}$\\
\textit{Faculty of Information Systems} \\
\textit{University of Information Technology - Vietnam National University} \\
Ho Chi Minh City, Vietnam \\
\{21521447@gm.uit.edu.vn, 21522049@gm.uit.edu.vn, nhutnm.17@grad.uit.edu.vn, thuannd@uit.edu.vn\}
}

\maketitle
\onecolumn
\begin{abstract}
This study addresses the challenges of data imbalance and feature bias in predicting autism spectrum disorder (ASD) using pediatric psychological records. We evaluate four classification models: Decision Tree, XGBoost, CatBoost, and Overall Local Accuracy (OLA) on real-world clinical data. Our findings highlight OLA's superior performance, achieved by mitigating feature dominance and effectively handling imbalanced datasets. By reducing reliance on specific data columns, OLA provides a balanced and robust approach to ASD classification, offering promising implications for early diagnosis in pediatric psychology.
\end{abstract}

% \begin{IEEEkeywords}
% [Từ khóa về chủ đề nghiên cứu mới]
% \end{IEEEkeywords}

\section{Introduction}
% Nêu bối cảnh, lý do chọn đề tài, các vấn đề đang tồn tại, và mục tiêu nghiên cứu của bài.
Autism Spectrum Disorder (ASD) is a complex neurodevelopmental condition marked by difficulties in social interaction, communication, and repetitive behaviors. Early and accurate diagnosis of ASD in children is vital for timely intervention, which can greatly improve long-term outcomes. However, applying clinical psychological records to predict ASD faces challenges, including data imbalance, feature bias, and limited integration of information technology (IT) with medical and psychological fields, which complicates obtaining reliable datasets. Effective data preprocessing-such as collecting, selecting, labeling, and refining data with guidance from psychology experts-is crucial to build a meaningful dataset for model training, ensuring it accurately reflects clinical realities.

@@

\section{Related Works}
% Tổng quan các công trình nghiên cứu liên quan, so sánh và xác định khoảng trống nghiên cứu.
Harshita Chandrappa, Swaathi BR, Prof. Swimpy Pahuja.~\cite{b1} proposed a hybrid machine learning framework for predicting Autism Spectrum Disorder (ASD) across all age groups. The system incorporates multiple ML classifiers such as Decision Tree, Naive Bayes, Random Forest, SVM, and Logistic Regression into a web-based diagnostic application. Through preprocessing and classification on a structured ASD dataset, their approach demonstrated high accuracy in predicting ASD traits. Among the models evaluated, Support Vector Machine and Random Forest yielded the best results in terms of accuracy and robustness. The authors highlighted the importance of early screening and presented a chatbot-based interface to assist users in understanding ASD symptoms and diagnosis.

José A. Sáez, Julián Luengo, Jerzy Stefanowski, Francisco Herrera ~\cite{b2} presented a novel approach for text classification that combines SMOTE-based oversampling with an IDF (Inverse Document Frequency) weighting mechanism to address class imbalance in textual datasets. The authors utilized a Deep Neural Network (DNN) as the main classifier and applied their method to benchmark datasets such as 20 Newsgroups. Their technique, referred to as SMOTE-IDF, effectively balances minority class samples while preserving the importance of rare but informative terms. Experimental results showed that the SMOTE-IDF+DNN model outperformed traditional oversampling techniques and baseline classifiers in terms of accuracy, precision, recall, and F1-score. The study emphasizes the importance of hybrid sampling and feature weighting in improving classification performance on imbalanced textual data.

Hanen Karamti, Raed Alharthi, Amira Al Anizi, Reemah M. Alhebshi, Ala’ Abdulmajid Eshmawi, Shtwai Alsubai, Muhammad Umer.~\cite{b3} introduced an enhanced oversampling technique named SMOTE-KNN, which improves upon the original SMOTE algorithm by integrating a K-Nearest Neighbors (KNN)-based adaptive mechanism. The proposed method dynamically selects appropriate neighbor samples for synthetic instance generation, aiming to reduce noise and overlap between classes in imbalanced datasets. The authors validated SMOTE-KNN on multiple benchmark datasets and compared its performance with traditional SMOTE, Borderline-SMOTE, and ADASYN. Experimental results demonstrated that SMOTE-KNN significantly enhances classification metrics such as precision, recall, and F1-score across various classifiers. The study underlines the importance of local data structure in oversampling and offers a more robust strategy for handling imbalance in binary classification problems.

@@

Youngkyu Hong, Eunho Yang.~\cite{b5} introduced a dual-branch learning framework that tackles bias in image classification tasks by separating biased and unbiased representations. Their method employs two key components: Bias-Contrastive Learning (BCL), which guides the biased branch to focus on spurious correlations, and Bias-Balanced Learning (BBL), which steers the unbiased branch to rely on robust features. By minimizing cross-entropy loss from the biased branch and maximizing agreement with ground truth labels in the unbiased branch, the framework effectively reduces model reliance on dataset-specific biases. Evaluations on benchmark datasets including Biased-MNIST and ImageNet-9 demonstrate that this method significantly outperforms conventional debiasing techniques, especially in scenarios where training labels are confounded by dominant visual cues.

Y. Zeng, J. Liu, H. Lam, and H. Namkoong.~\cite{b7} investigated the use of Large Language Model (LLM) embeddings to enhance the robustness of tabular classifiers under distributional shifts. Their method integrates LLM-derived feature representations with original tabular features and applies group-wise robust optimization at test time to adapt to both covariate (X) and label (Y) shifts. The study shows that this approach improves the generalization performance of traditional classifiers such as XGBoost and logistic regression across diverse datasets. The results emphasize the potential of LLM embeddings as transferable features for tabular classification tasks under domain shift scenarios.

@@

\subsubsection{Private Dataset}
The private dataset was collected from the medical records of the Psychology Clinic, featuring a more complex structure and class imbalance due to medical records in favor of disease data. It contains over 600 samples, with richer information such as personal information, acknowledging behavior from relatives of children and evaluating from a specialist after a direct examination at the clinic. A major challenge in this dataset is the excessive dominance of a single feature, which represents the standard assessment by doctors. This feature exhibits a bias tendency, affecting the final clinical outcomes and resulting in model decisions with high experimental accuracy but prone to misapplication in practical settings.\\

a, Raw data: This study utilizes a dataset comprising over \textbf{1,200 pediatric psychological records} collected from a child psychology clinic, covering various conditions such as Autism Spectrum Disorder (ASD), intellectual developmental disorder, language disorder, and attention deficit hyperactivity disorder. The initial step involved standardizing the data by converting PDF records into structured Excel tables. The records were then categorized by condition, resulting in the selection of over 800 cases related to ASD and non-clinical cases.\\

b, Filter data: Records with missing or substandard data were filtered out, yielding a refined dataset of \textbf{about 600 records}. Then, data labeling was performed to assign features known to influence ASD prediction, including eye contact~\cite{b8}, pointing gestures~\cite{b9}, response to name~\cite{b10}, joint attention~\cite{b11}, imitation~\cite{b12}, and symbolic play~\cite{b13}... These behavioral indicators were selected based on their strong association with early autism markers and clinical diagnostic standards. To ensure privacy, personal information was anonymized through encoding before further preprocessing.\\

c, Dataset prepared:
\begin{itemize}
    \item \textbf{Number of samples:} 594 records
    \item \textbf{Class distribution:} 550/594 records ASD-Tracking and ASD about 93\%
    \item \textbf{Number and types of features:} 36 columns included raw columns and label columns
\end{itemize}
\subsection{Data Preprocessing}
\textbf{a, Overview Data Preprocessing}
\begin{table}[H]
\caption{Data Preprocessing Steps for private dataset}
\label{tab:preprocessing}
\centering
\resizebox{\columnwidth}{!}{
\begin{tabular}{|p{2.2cm}|p{7.3cm}|}
\hline

@@

    \begin{figure}[H]
    \centering
    \includegraphics[width=0.9\linewidth]{images/decisionTree.png}
    \caption{Decision Tree structure}
    \label{fig:decision-tree}
    \end{figure}
    \item \textbf{XGBoost} \\
    An ensemble learning method based on gradient boosting, combines multiple weak learners (decision trees) to improve predictive performance. It handles complex feature interactions effectively and includes regularization to mitigate overfitting. Despite its robustness, XGBoost can still exhibit bias toward dominant features in imbalanced datasets, limiting its generalizability in pediatric ASD classification.
    \begin{figure}[H]
    \centering
    \includegraphics[width=0.9\linewidth]{images/XGB.png}
    \caption{XGBoosting structure}
    \label{fig:xgboost-structure}
    \end{figure}

    \item \textbf{CatBoost} \\
    An advanced Gradient Boosting Decision Trees (GBDT) algorithm, builds an ensemble of symmetric (oblivious) decision trees to enhance predictive performance and computational efficiency. Unlike traditional Decision Trees, which rely on a single, overfitting-prone model, CatBoost combines multiple weak learners with ordered boosting and regularization to reduce overfitting, achieving superior generalization on complex datasets. It was configured with a depth of 6 and 100 iterations, leveraging its ability to process ordinal features (e.g., poor-moderate-good encodings) directly.

    Compared to Decision Trees, CatBoost excels in handling categorical features through automated label encoding with Bayesian smoothing, eliminating the need for manual preprocessing. Relative to XGBoost, CatBoost offers improved categorical feature processing, optimized GPU performance via histogram-based computations, and automatic feature combinations for complex data patterns. Its structure, illustrates the ensemble of oblivious trees and the ordered boosting process, highlighting efficient memory usage and parallelization.

    \begin{figure}[H]
    \centering
    \includegraphics[width=0.9\linewidth]{images/Catboost.png}
    \caption{CatBoost Algorithm Structure}
    \label{fig:catboost-structure}
    \end{figure}

@@

    \item \textbf{OLA (Online Local Accuracy \cite{b16})} \\
    a, Overview OLA model\\

    A Dynamic Classifier Selection (DCS) technique from the DESlib library~\cite{b17}, dynamically selects the most competent classifier for each test sample based on its local accuracy within a region of competence. For each test instance $x$, OLA identifies the $k$-nearest neighbors in the training set, evaluates the classification accuracy of each base classifier on these neighbors, and selects the classifier with the highest local accuracy to predict $x$.

@@

Implemented with custom code to integrate with the SMOTE-IPF preprocessed dataset, OLA mitigates feature bias and improves robustness on imbalanced data. OLA was hypothesized to outperform the other methods due to its ability to adapt to local data patterns, making it a promising tool for accurate and fair ASD diagnosis in pediatric settings.\\

b, Reducing Feature Bias Impact with OLA
\begin{table}[H]
\caption{Enhanced OLA Implementation Pipeline}
\label{tab:enhanced-ola}
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{|p{3.8cm}|p{5.5cm}|p{5.8cm}|}
\hline
\textbf{Step} & \textbf{Objective} & \textbf{Actions and Outcome} \\
\hline
\rule{0pt}{15pt} \textbf{Step 1: Data Normalization and Dimensionality Reduction with PCA} & Reduce correlations between features and mitigate the impact of biased features like \textit{TiepXucMat}. & 
\begin{itemize}
    \item Apply StandardScaler to normalize data (mean=0, variance=1).
    \item Use PCA to retain 95\% variance and minimize dependence on \textit{TiepXucMat}.
    \item \textbf{Outcome:} Bias from dominant features is minimized while essential information is preserved.
\end{itemize} \\
\hline

\rule{0pt}{15pt} \textbf{Step 2: Create a Diverse Classifier Pool} & Build a diverse set of classification models to enhance OLA’s flexibility and accuracy. & 
\begin{itemize}
    \item Model Group 1 (PCA-based): DecisionTreeClassifier, XGBClassifier.
    \item Model Group 2 (excluding \textit{TiepXucMat}): RandomForestClassifier.
    \item Model Group 3 (raw data): KNeighborsClassifier.
    \item \textbf{Outcome:} Enables OLA to select the most appropriate model per sample.
\end{itemize} \\
\hline

\rule{0pt}{15pt} \textbf{Step 3: Train OLA on PCA-based Models} & Use OLA to dynamically select the best classifier for each test sample based on local accuracy. & 
\begin{itemize}
    \item Train OLA on PCA-based classifiers using kNN ($k=7$).
    \item Compute local accuracy in each neighborhood.
    \item \textbf{Outcome:} Classifier is chosen based on local performance, reducing global feature bias.
\end{itemize} \\
\hline

\rule{0pt}{15pt} \textbf{Step 4: Combine Prediction Probabilities from All Model Groups} & Enhance prediction accuracy and robustness by integrating outputs from diverse model types. & 
\begin{itemize}
    \item Aggregate probabilities from all groups (PCA, reduced, raw).
    \item Average the results to improve ROC AUC.
    \item \textbf{Outcome:} Increases reliability by mitigating influence of any single biased model group.
\end{itemize} \\
\hline
\end{tabular}}
\end{table}

\end{enumerate}

@@

\subsubsection{Evaluation Metrics}

\paragraph{Accuracy}

Measures the proportion of correctly classified samples out of the total.

\textbf{Formula}:
\[
\text{Accuracy} = \frac{TP + TN}{TP + TN + FP + FN}
\]

\textbf{Explanation}:
\begin{itemize}
    \item $TP$: True Positives – correctly predicted ASD cases.
    \item $TN$: True Negatives – correctly predicted non-ASD cases.
    \item $FP$: False Positives – non-ASD cases predicted as ASD.
    \item $FN$: False Negatives – ASD cases predicted as non-ASD.
\end{itemize}

\textbf{Significance}: Accuracy gives a general performance measure but can be misleading on imbalanced datasets like ASD.

\textbf{Evaluation}: Values closer to 1 are better, but should be interpreted cautiously in imbalanced settings.

\paragraph{Recall (Sensitivity)}

Measures the proportion of actual ASD cases correctly identified.

\textbf{Formula}:
\[
\text{Recall} = \frac{TP}{TP + FN}
\]

\textbf{Significance}: Critical for medical applications, ensuring ASD cases are not missed.

\textbf{Evaluation}: High recall reduces the risk of undetected ASD diagnoses.

\paragraph{Specificity}

Measures the proportion of non-ASD cases correctly identified.

\textbf{Formula}:
\[
\text{Specificity} = \frac{TN}{TN + FP}
\]

\textbf{Significance}: Important for reducing false alarms (false positives).

\textbf{Evaluation}: High specificity prevents unnecessary concern or intervention for non-ASD individuals.

\paragraph{Precision}

Measures the proportion of predicted ASD cases that are correct.

\textbf{Formula}:
\[
\text{Precision} = \frac{TP}{TP + FP}
\]

\textbf{Significance}: Indicates reliability of positive (ASD) predictions.

\textbf{Evaluation}: High precision reduces false ASD labels.

\paragraph{F1-score}

The harmonic mean of precision and recall.

\textbf{Formula}:
\[
\text{F1-score} = 2 \cdot \frac{\text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}}
\]

\textbf{Significance}: Provides a balanced measure for imbalanced data, combining both false positives and false negatives.

\paragraph{ROC AUC}

Area under the Receiver Operating Characteristic curve.

\textbf{Formula}:
\begin{itemize}
    \item True Positive Rate (Recall): 
    \[
    \text{TPR} = \frac{TP}{TP + FN}
    \]
    \item False Positive Rate (FPR): 
    \[
    \text{FPR} = \frac{FP}{FP + TN}
    \]
\end{itemize}

\textbf{Significance}: Reflects the model’s ability to distinguish between ASD and non-ASD classes across thresholds.

\textbf{Evaluation}: Ranges from 0.5 (random) to 1.0 (perfect).

\paragraph{G-Mean}

Geometric mean of recall and specificity.

\textbf{Formula}:
\[
\text{G-Mean} = \sqrt{\text{Recall} \cdot \text{Specificity}}
\]

\textbf{Significance}: Balances performance across classes by combining recall and specificity, addressing the limitations of accuracy in imbalanced datasets like ASD, where minority class detection (e.g., ASD cases) is critical.

\textbf{Evaluation}: Ranges from 0 to 1; a value close to 1 indicates balanced and effective classification across both classes, while lower values suggest imbalance or poor performance, necessitating model refinement.

@@

@@

\paragraph{\textbf{Models}}
\begin{itemize}
\item \textbf{Baseline Models:} Decision Tree, XGBoost, CatBoost, and the initial OLA configuration.
\item \textbf{Enhanced OLA:} After identifying bias in the private dataset, OLA was refined to include multiple classifier pools to enhance robustness and fairness.
\end{itemize}

\subsection{Experimental Workflow}
The experimental pipeline consisted of three main steps:

\paragraph{\textbf{Step 1: Initial Evaluation on Both Datasets}}

@@

\paragraph{\textbf{Step 2: Analysis of features Bias on Private Dataset}}\\

@@

\paragraph{\textbf{Step 3: Enhanced OLA with Feature Bias Mitigation}}\\

Mitigate the influence of \textit{TiepXucMat} and improve generalization in OLA.\\
\textbf{Process:} The OLA classifier pool was enhanced with:
\begin{itemize}
\item PCA-based models (to decorrelate features),
\item Models excluding \textit{TiepXucMat} entirely (to enforce fairness),
\item KNN-based models on raw and reduced data.
\end{itemize}
This strategy helped reduce model dependency on a single dominant feature and increased fairness in detecting ASD across varied behavioral patterns.

\onecolumn
\subsection{Experimental Results}
This subsection presents the results of the three experimental steps, including performance metrics for all models on both datasets, feature importance analysis, and the impact of the enhanced OLA model. The results are reported using single-run evaluations, 5-fold cross-validation, and visualizations.

>>>>

\documentclass[conference]{IEEEtran}

% ---------- PACKAGES ----------
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{lipsum}
\usepackage{url}
\usepackage{cite}
\usepackage{url}

% ---------- TABLE ADJUSTMENT ----------
\usepackage{adjustbox}
\usepackage{longtable}  % Nếu cần bảng nhiều dòng
\usepackage{stfloats}   % Cho phép table* nằm dưới cuối trang

% ---------- BEGIN DOCUMENT ----------
\begin{document}

% ---------- TITLE ----------
\title{Enhanced Overall Local Accuracy Framework For Addressing Feature Bias and Data Imbalance in ASD Classification}

\author{
Hao Nguyen Thi Bich $^{1}$, Thanh Nguyen Van Quoc $^{2}$, Nhut Nguyen Minh$^{3}$, Thuan Nguyen Dinh$^{4}$\\
\textit{Faculty of Information Systems} \\
\textit{University of Information Technology - Vietnam National University} \\
Ho Chi Minh City, Vietnam \\
\{ 21522049@gm.uit.edu.vn, 21521447@gm.uit.edu.vn, nhutnm.17@grad.uit.edu.vn, thuannd@uit.edu.vn\}
}

% \author{
%     \IEEEauthorblockN{Author One, Author Two}
%     \IEEEauthorblockA{
%         Department, University \\
%         Email: \{author1, author2\}@univ.edu}
% }

\maketitle
\IEEEpeerreviewmaketitle

% ---------- ABSTRACT ----------
\begin{abstract}
This study addresses the challenges of data imbalance and feature bias in predicting autism spectrum disorder (ASD) using pediatric psychological records. We evaluate four classification models: Decision Tree, XGBoost, CatBoost, and Overall Local Accuracy (OLA) on real-world clinical data. Our findings highlight OLA's superior performance, achieved by mitigating feature dominance and effectively handling imbalanced datasets. By reducing reliance on specific data columns, OLA provides a balanced and robust approach to ASD classification, offering promising implications for early diagnosis in pediatric psychology.
\end{abstract}

% ---------- KEYWORDS ----------
\begin{IEEEkeywords}
ASD classification, OLA framework, Decision Tree, XGBoost, CatBoost, SMOTE-IPF
\end{IEEEkeywords}
\section{Introduction}

Autism Spectrum Disorder (ASD) is a complex neurodevelopmental condition marked by difficulties in social interaction, communication, and repetitive behaviors. Early and accurate diagnosis of ASD in children is vital for timely intervention, which can greatly improve long-term outcomes. However, applying clinical psychological records to predict ASD faces challenges, including data imbalance, feature bias, and limited integration of information technology (IT) with medical and psychological fields, which complicates obtaining reliable datasets. Effective data preprocessing-such as collecting, selecting, labeling, and refining data with guidance from psychology experts-is crucial to build a meaningful dataset for model training, ensuring it accurately reflects clinical realities.

@@

\section{Related Works}

% Tổng quan các công trình nghiên cứu liên quan, so sánh và xác định khoảng trống nghiên cứu.
Harshita Chandrappa, Swaathi BR, Prof. Swimpy Pahuja.~\cite{b1} proposed a hybrid machine learning framework for predicting Autism Spectrum Disorder (ASD) across all age groups. The system incorporates multiple ML classifiers such as Decision Tree, Naive Bayes, Random Forest, SVM, and Logistic Regression into a web-based diagnostic application. Through preprocessing and classification on a structured ASD dataset, their approach demonstrated high accuracy in predicting ASD traits. Among the models evaluated, Support Vector Machine and Random Forest yielded the best results in terms of accuracy and robustness. The authors highlighted the importance of early screening and presented a chatbot-based interface to assist users in understanding ASD symptoms and diagnosis.

José A. Sáez, Julián Luengo, Jerzy Stefanowski, Francisco Herrera ~\cite{b2} introduced SMOTE-IDF, a novel text classification method combining SMOTE oversampling with IDF weighting to tackle class imbalance in textual datasets. Using a Deep Neural Network (DNN) classifier, SMOTE-IDF balances minority class samples while preserving the significance of rare, informative terms. Tested on datasets like 20 Newsgroups, it outperformed traditional oversampling and baseline classifiers in accuracy, precision, recall, and F1-score, highlighting the value of hybrid sampling and feature weighting for imbalanced textual data.

% Hanen Karamti, Raed Alharthi, Amira Al Anizi, Reemah M. Alhebshi, Ala’ Abdulmajid Eshmawi, Shtwai Alsubai, Muhammad Umer.~\cite{b3} introduced an enhanced oversampling technique named SMOTE-KNN, which improves upon the original SMOTE algorithm by integrating a K-Nearest Neighbors (KNN)-based adaptive mechanism. The proposed method dynamically selects appropriate neighbor samples for synthetic instance generation, aiming to reduce noise and overlap between classes in imbalanced datasets. The authors validated SMOTE-KNN on multiple benchmark datasets and compared its performance with traditional SMOTE, Borderline-SMOTE, and ADASYN. Experimental results demonstrated that SMOTE-KNN significantly enhances classification metrics such as precision, recall, and F1-score across various classifiers. The study underlines the importance of local data structure in oversampling and offers a more robust strategy for handling imbalance in binary classification problems.

@@

Youngkyu Hong, Eunho Yang.~\cite{b5} proposed a dual-branch learning framework to address bias in image classification. It uses Bias-Contrastive Learning (BCL) to capture spurious correlations in a biased branch and Bias-Balanced Learning (BBL) to focus on robust features in an unbiased branch. By minimizing biased branch loss and maximizing unbiased branch agreement with true labels, the method reduces reliance on dataset biases, outperforming traditional debiasing techniques on datasets like Biased-MNIST and ImageNet-9.

Y. Zeng, J. Liu, H. Lam, and H. Namkoong.~\cite{b7} explored using Large Language Model (LLM) embeddings to improve the robustness of tabular classifiers like XGBoost and logistic regression under distributional shifts. By combining LLM-derived features with original tabular features and applying group-wise robust optimization at test time, their method enhances generalization performance across diverse datasets, effectively handling covariate and label shifts.

@@

\subsubsection{Private Dataset}
The private dataset was collected from the medical records of the Psychology Clinic, featuring a more complex structure and class imbalance due to medical records in favor of disease data. It contains over 600 samples, with richer information such as personal information, acknowledging behavior from relatives of children and evaluating from a specialist after a direct examination at the clinic. A major challenge in this dataset is the excessive dominance of a single feature, which represents the standard assessment by doctors. This feature exhibits a bias tendency, affecting the final clinical outcomes and resulting in model decisions with high experimental accuracy but prone to misapplication in practical settings.\\
\textbf{a, Raw data}: This study utilizes a dataset comprising over \textbf{1,200 pediatric psychological records} collected from a child psychology clinic, covering various conditions such as Autism Spectrum Disorder (ASD), intellectual developmental disorder, language disorder, and attention deficit hyperactivity disorder. The initial step involved standardizing the data by converting PDF records into structured Excel tables. The records were then categorized by condition, resulting in the selection of over 800 cases related to ASD and non-clinical cases.\\
\textbf{b, Filter data}: Records with missing or substandard data were filtered out, yielding a refined dataset of \textbf{about 600 records}. Then, data labeling was performed to assign features known to influence ASD prediction, including eye contact~\cite{b8}, pointing gestures~\cite{b9}, response to name~\cite{b10}, joint attention~\cite{b11}, imitation~\cite{b12}, and symbolic play~\cite{b13}... These behavioral indicators were selected based on their strong association with early autism markers and clinical diagnostic standards. To ensure privacy, personal information was anonymized through encoding before further preprocessing.\\
\textbf{c, Dataset prepared}:
\begin{itemize}
    \item \textbf{Number of samples:} 594 records
    \item \textbf{Class distribution:} 550/594 records ASD-Tracking and ASD about 93\%
    \item \textbf{Number and types of features:} 36 columns included raw columns and label columns
\end{itemize}
\subsection{Data Preprocessing}
\textbf{a, Overview Data Preprocessing}
\begin{table}[H]
\caption{Data Preprocessing Steps for private dataset}
\label{tab:preprocessing}
\centering
\resizebox{\columnwidth}{!}{
\begin{tabular}{|p{2.2cm}|p{7.3cm}|}
\hline

@@

    % \begin{figure}[H]
    % \centering
    % \includegraphics[width=0.9\linewidth]{images/decisionTree.png}
    % \caption{Decision Tree structure}
    % \label{fig:decision-tree}
    % \end{figure}
    \item \textbf{XGBoost} \\
    An ensemble learning method based on gradient boosting, combines multiple weak learners (decision trees) to improve predictive performance. It handles complex feature interactions effectively and includes regularization to mitigate overfitting. Despite its robustness, XGBoost can still exhibit bias toward dominant features in imbalanced datasets, limiting its generalizability in pediatric ASD classification.
    % \begin{figure}[H]
    % \centering
    % \includegraphics[width=0.9\linewidth]{images/XGB.png}
    % \caption{XGBoosting structure}
    % \label{fig:xgboost-structure}
    % \end{figure}

    \item \textbf{CatBoost} \\
    An advanced Gradient Boosting Decision Trees (GBDT) algorithm, builds an ensemble of symmetric (oblivious) decision trees to enhance predictive performance and computational efficiency. Unlike traditional Decision Trees, which rely on a single, overfitting-prone model, CatBoost combines multiple weak learners with ordered boosting and regularization to reduce overfitting, achieving superior generalization on complex datasets. It was configured with a depth of 6 and 100 iterations, leveraging its ability to process ordinal features (e.g., poor-moderate-good encodings) directly.
    Compared to Decision Trees, CatBoost excels in handling categorical features through automated label encoding with Bayesian smoothing, eliminating the need for manual preprocessing. Relative to XGBoost, CatBoost offers improved categorical feature processing, optimized GPU performance via histogram-based computations, and automatic feature combinations for complex data patterns. Its structure, illustrates the ensemble of oblivious trees and the ordered boosting process, highlighting efficient memory usage and parallelization.

    % \begin{figure}[H]
    % \centering
    % \includegraphics[width=0.9\linewidth]{images/Catboost.png}
    % \caption{CatBoost Algorithm Structure}
    % \label{fig:catboost-structure}
    % \end{figure}

@@

    \item \textbf{OLA (Online Local Accuracy \cite{b16})} \\
    \textbf{a, Overview OLA model}\\
A Dynamic Classifier Selection (DCS) technique from the DESlib library~\cite{b17}, dynamically selects the most competent classifier for each test sample based on its local accuracy within a region of competence. For each test instance $x$, OLA identifies the $k$-nearest neighbors in the training set, evaluates the classification accuracy of each base classifier on these neighbors, and selects the classifier with the highest local accuracy to predict $x$.

@@

Implemented with custom code to integrate with the SMOTE-IPF preprocessed dataset, OLA mitigates feature bias and improves robustness on imbalanced data. OLA was hypothesized to outperform the other methods due to its ability to adapt to local data patterns, making it a promising tool for accurate and fair ASD diagnosis in pediatric settings.\\
\textbf{b, Enhancing Feature Bias Impact with OLA}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.95\linewidth]{images/structure_imp_ola.png} 
    \caption{Enhanced OLA Pipeline}
    \label{fig:ola_pipeline}
\end{figure}
\begin{itemize}
\item \textbf{Step 1: Data Normalization and Dimensionality Reduction with PCA}~\\
\textbf{Description:} This step standardizes and reduces the dimensionality of the dataset to minimize the dominance of biased features such as \textit{TiepXucMat}. StandardScaler is applied (mean=0, variance=1), followed by PCA with $n\_components=0.95$ to retain 95\% of variance.\\
\textbf{Output:} Normalized and PCA-transformed data.\\
\textbf{Purpose:} Reduces noise and bias from dominant features while preserving essential structure.

\item \textbf{Step 2: Create 3 Diverse Classifier Pool}~\\
\textbf{Description:} A diverse set of classifiers is built to enhance flexibility and generalization:\\
\textit{Group 1 (PCA-based)}: DecisionTreeClassifier, XGBClassifier trained on PCA data.\\
\textit{Group 2 (Bias-controlled)}: RandomForestClassifier trained on raw data excluding \textit{TiepXucMat}.\\
\textit{Group 3 (Raw data)}: KNeighborsClassifier ($k=5$) trained directly on raw data.\\
\textbf{Output:} \texttt{pool\_classifiers} containing all trained models.\\
\textbf{Purpose:} Promotes diversity, allowing OLA to adaptively choose the most suitable model.

\item \textbf{Step 3: Local Accuracy Estimation and Training OLA Model with Group 1}~\\
\textbf{Description:} OLA is trained using models in Group 1 and estimates the local accuracy of each classifier to dynamically select the best performer for each input sample.\\
\textbf{Output:} Trained OLA model using PCA-transformed data and \texttt{ola\_pool} (DecisionTreeClassifier and XGBClassifier).\\
\textbf{Purpose:} Improves individual prediction decisions by selecting models with highest local accuracy.

\item \textbf{Step 4: Prediction Probabilities from All Model Groups}~\\
\textbf{Description:} Final predictions are made using OLA and ensemble probabilities from all three groups:\\
- OLA produces label predictions ($y\_pred$) using \texttt{ola\_pool}.\\
- All models generate probabilities via \texttt{predict\_proba}; PCA-based models predict on PCA-transformed data, while others use raw data.\\
- The probabilities are averaged into a final $y\_prob$ score.\\
\textbf{Output:} Final prediction labels ($y\_pred$) and ensemble probabilities ($y\_prob$).\\
\textbf{Purpose:} Increases reliability by mitigating influence of any single biased model group, useful for metrics like ROC AUC.
\end{itemize}
\end{enumerate}

@@

\subsubsection{Evaluation Metrics}: Accuracy, Recall, Specificity, Precision, F1-score, ROC-AUC, G-Mean

@@

% ---------- CONTINUED SECTIONS ----------

@@

\paragraph{\textbf{Models}}
\begin{itemize}
\item \textbf{Baseline Models:} Decision Tree, XGBoost, CatBoost, and the initial OLA configuration.
\item \textbf{Enhanced OLA:} After identifying bias in the private dataset, OLA was refined to include multiple classifier pools to enhance robustness and fairness.\\

\end{itemize}

\subsection{Experimental Workflow}
The experimental pipeline consisted of three main steps:
\paragraph{\textbf{Step 1: Balancing Private Dataset}}

SMOTE-KNN and SMOTE-IPF were evaluated to generate synthetic samples for the minority class (label 0), ensuring a balanced distribution aligned with PCA components for enhanced practical relevance.\\
\textbf{Process:} The private dataset, initially comprising over 600 records, exhibited significant target label imbalance, with a skew toward ASD cases (label 1).

\paragraph{\textbf{Step 2: Initial Evaluation on Both Datasets}}

@@

\paragraph{\textbf{Step 3: Analysis of features Bias on Private Dataset}}\\

@@

\paragraph{\textbf{Step 4: Enhanced OLA with Feature Bias Mitigation}}\\

Mitigate the influence of \textit{TiepXucMat} and improve generalization in OLA.\\
\textbf{Process:} The OLA classifier pool was enhanced with:
\begin{itemize}
\item PCA-based models (to decorrelate features),
\item Models excluding \textit{TiepXucMat} entirely (to enforce fairness),
\item KNN-based models on raw and reduced data.
\end{itemize}
This strategy helped reduce model dependency on a single dominant feature and increased fairness in detecting ASD across varied behavioral patterns.
\subsection{Experimental Results}
This subsection presents the results of the four experimental steps, including performance metrics for all models on both datasets, feature importance analysis, and the impact of the enhanced OLA model. The results are reported using single-run evaluations, 5-fold cross-validation, and visualizations.

\subsubsection{Evaluation Balancing Private Dataset - Step 1}
Two chart 2D PCA visualization bellow reveal that SMOTE-IPF yields a more uniform distribution compared to SMOTE-KNN, with synthetic samples exhibiting tighter clustering and reduced overlap between classes. This indicates SMOTE-IPF’s higher reliability in balancing the private dataset while preserving class separability, enhancing the robustness of subsequent model training.\\
Therefore, private dataset with SMOTE-IPF over 800 records were balanced used for model.

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.48\linewidth}
        \includegraphics[width=\linewidth]{images/pca_smote_knn.png}
        \caption{PCA with SMOTE-KNN}
        \label{fig:pca-smote-knn}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.48\linewidth}
        \includegraphics[width=\linewidth]{images/pca_smote_ipf.png}
        \caption{PCA with SMOTE-IPF}
        \label{fig:pca-smote-ipf}
    \end{subfigure}
    \caption{Comparison of PCA visualizations on private dataset using SMOTE-KNN and SMOTE-IPF.}
    \label{fig:pca-comparison}
\end{figure}
